An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies

### 2017-05-12 main.tex
Add Owen C as author.
Write a first draft.
This has removed a number of things to make the paper a bit more
efficient. I have taken Marc's text about training strategies and moved
it around to fit better. It is also one of the main narratives of the
story now.

Note that I've removed "we" and "our" throughout.

% Title
\title{An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies}
\author{Marc Harper \and Vincent Knight \and Nikoleta E. Glynatsi} % TODO Authors?
\date{}

@@

\begin{abstract}
    The Iterated Prisoner's Dilemma is a well established framework for
    the study of emergent behaviour. In this paper an extensive numerical
    study of the evolutionary dynamics of this framework are presented. Fixation
    probabilities for Moran processes are obtained for 164
    different strategies. We find that players with long memories and
    sophisticated behaviours outperform many strategies that perform well
    in a two player setting.
\end{abstract}

@@

The Prisoner's Dilemma (PD)~\cite{Flood1958} is a fundamental two player game
used to model a large variety of strategic interactions. Each player can choose
between cooperation (C) or defection (D). The decisions are made simultaneously
and independently. The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This extended form is called the Iterated Prisoner's
Dilemma (IPD).

The Moran Process \cite{Moran1957} is a model of
evolutionary population dynamics that has been used to gain insights about the
evolutionary stability and fixation of strategies in a number of settings.
Similarly since the first Iterated Prisoner's Dilemma (IPD) tournament described in
\cite{Axelrod1980a}, the Prisoner's dilemma has been used to understand the
evolution of cooperative behaviour in complex systems. Several earlier
works have studied iterated games in the context of the prisoner's dilemma
\cite{Nowak, Stewart2012}.

This manuscript provides a detailed numerical analysis of \textbf{164} complex
and adaptive strategies for the IPD\@. This is made possible by the Axelrod
library \cite{axelrodproject}, an effort to provide software for reproducible
research for the IPD\@. The library now contains over 150 parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. The library can conduct matches, tournaments and population
dynamics with variations including noise and spatial structure.

We present fixation probabilities for all pairs of
strategies in the library, identifying those are effective invaders and
those resistant to invasion, for population sizes $N=2$ to $N=14$.

In particular we address the following questions:
\begin{enumerate}
    \item What strategies are good invaders?
    \item What strategies are good at resisting invasion?
    \item How does the population size affect these findings?
\end{enumerate}

While our results agree with some of the published literature, we find that:

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Long memory strategies are generally more effective than short
       memory strategies
 \item Complex strategies can be effective
\end{enumerate}

\subsection{The Moran Process}

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process, a stochastic birth death process on a finite population
in which the population size stays constant over time. Individuals are
\textbf{selected} according to a given fitness landscape. Once selected, the
individual is reproduced and similarly another individual is chosen to be
removed from the population. In some settings mutation is also considered but
without mutation (the case considered in this work) this process will arrive at
an absorbing state where the population is entirely made up of players of
one strategy. The probability with which a given strategy is the survivor is
called the fixation probability. A more detailed analytic description of this
is given in Section~\ref{sec:validation}.

@@

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider non-sophisticated strategies. Some work has
looked at evolutionary stability of strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely-used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies.  The subset included memory one strategies, strategies that recall
the events of the previous round only.

Of particular interest are the zero determinant strategies introduced in
\cite{Press2012} and highly-praised \cite{Stewart2012} it was argued that
generous ZD strategies are robust against invading strategies. However, in
\cite{Lee2015} a strategy using machine learning techniques was capable of
resisting invasion and also able to invade any memory one strategy.  Recent work
\cite{Hilbe2017} has investigated the effect of memory length on strategy
performance and the emergence of cooperation but this is not done in Moran
process context and only considers specific cases of memory 2 strategies.  In
\cite{Adami2013} it was recognized that many zero determinant strategies do not
fare well against themselves. This is a disadvantage for the Moran process where
the best strategies cooperate well with other players using the same strategy.

The contribution of this work is a detailed and extensive analysis of absorption
probabilities for 164 strategies. These strategies and the numerical simulations
are from~\cite{axelrodproject} which is an open source research library written
for the study of the IPD\@. The strategies and simulation frameworks are
automatically tested to an extraordinarily high degree of coverage in accordance
with best research software practices. The large
number of strategies are available thanks to the open source nature of the
project with over 40 contributions made by different programmers and researchers.

Section~\ref{sec:methodology} will explain the methodological approach used,
Section~\ref{sec:validation} will validate the methodology by comparing
simulated results to analytical results in some cases. The main results of this
manuscript are presented in Section~\ref{sec:empirical_results} which will
present a detailed analysis of all the data generated. Finally,
Section~\ref{sec:conclusion} will conclude and offer future avenues for the work
presented here.

\section{Methodology}\label{sec:methodology}

To carry out this large numerical experiment, 164 strategies are used from the
axelrod library: \cite{axelrodproject}.

The axelrod library contains many machine learning based strategies trained
with reinforcement learning algorithms. This work includes three additional
strategies trained specifically to excel at the Moran process.
Appendix~\ref{app:list_of_players} shows all the players in question. More
information about each player can be obtained in the documentation for
\cite{axelrodproject}. There are \input{./tex/num_stochastic.tex}stochastic and
\input{./tex/num_deterministic.tex}deterministic strategies. Their memory
depth, defined by the number of rounds of history used by the strategy each round,
is shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if
the strategy uses the entire history of play (whatever its length). For example,
a strategy that utilizes a handshaking mechanism where the opponents actions
on the first few rounds of play determines the strategies subsequent behavior
would have infinite memory depth.

All of the
strategies are trained with an evolutionary algorithm that perturbs strategy
% TODO Add a generic text book reference
parameters and optimizes an objective function, such as the mean total score
against all other opponents or the mean fixation probability over a large
number of Moran processes. The objective functions can include standard
variations such as noisy matches, varying the number of rounds of play, and the
other typical parameters used for the IPD. Variation is introduced via mutation
and crossover of parameters, and the best performing strategies are carried to
the next generation along with new variants. Similar methods appear in the
literature \cite{Ashlock2006}.

All of our training code is available in the
Axelrod repository with documentation to train new strategies easily. Training
% TODO Include link etc
typically takes less than 200 iterations and can be completed within several
hours on commodity hardware.  The three strategies trained specifically for
this study are Trained FSM 1, 2, and 3 (TF1 - TF3), based on finite state
machines of 8, 16, and 16 states respectively (see Figures \ref{}).
% TODO Include images of FSMs
These strategies were trained with the objective function of mean fixation
probabilities for Moran processes starting at initial population states
consisting of \(N/2\) individuals of the training candidates and \(N/2\)
individuals of an opponent strategy, for all of the short run time strategies
in the library (approximately 150 opponents at the time of training):

@@

Each matchup of players was run to fixation for the specified number of
repetitions to estimate the absorption probabilities. The trained algorithms
were run for less than 50 generations.

TF1 cooperates and defects with
various cycles depending on the opponent's actions. TF1 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective, it was selected for inclusion
because of the lack of an apparent handshake.

TF2 always starts
with CD and will defect against opponents that start with DD. It plays CDD
against itself and then cooperates thereafter. There is a longer complex
handshake which eventually results in mutual cooperation with Firm but Fair,
Fortress3, Fortress4, and Grofman (always) and Evolved HMM 5 and GTFT
(depending on the random seed).

@@

For both TF2 and TF3 a handshake
mechanism naturally emerges from the structure of the underlying finite state
machine. This behavior is an outcome of the evolutionary process and is in no
way hard-coded or included via an additional mechanism. Some of the existing
strategies in the library were trained on earlier versions of the library
(smaller total opponents used in training) based on neural networks, lookup
tables and stochastic variants trained with a particle swarm algorithm, finite
state machines and a stochastic variant mimicking hidden Markov models, and
neural networks. These strategies were trained to win IPD tournaments using an
objective function that computes the mean match payoffs against a collection of
opponents. Detailed descriptions and the performance of these strategies in IPD
tournaments will be described in another manuscript.
% TODO Think of how to reference this upcoming manuscript

@@

Each strategy pair is run for 1000 runs of the Moran process to fixation
with starting population distributions of $(1, N-1)$,
$(N/2, N/2)$ and $(N-1 , 1)$, for \(N\) from 2 through 14. The
fixation probability is then empirically computed for each combination of
starting distribution and value of \(N\).

Our software can carry out exact simulations of the Moran process. Since some of
the strategies have a high computational cost or are stochastic, we sample from
a large number of match outcomes for the pairs of players for use in computing
fitnesses in the Moran process. This approach was verified to agree with
unsampled calculations to a high degree of accuracy in specific cases.
% TODO Perhaps write some pseudo code for this?

Figure~\ref{fig:number_of_stochastic_match_outcomes} shows the distribution of
the number of outcomes between all strategy pairs.

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{./img/number_of_match_outcomes.pdf}
        \caption{All matches}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{./img/number_of_stochastic_match_outcomes.pdf}
        \caption{Stochastic matches}
    \end{subfigure}%
    \caption{The distribution of the number of unique outcomes used as the
    cached results}
    \label{fig:number_of_stochastic_match_outcomes}
\end{figure}

Section~\ref{sec:validation} will validate the methodology used here against
known theoretic results.

@@

Using this it is a known result that the fixation probability of the first
strategy in a population of \(i\) individuals of the first type (and \(N-i\)
individuals of the second. We have:

@@

A neutral strategy will have fixation probability $x_i = i/N$. Alternatively,
we can frame the outcomes in terms of relative fitness. For a strategy
of relative fitness $r$ the fixation probability is well-known to be

\[ x_i = \frac{1 - r^{-i}}{1 - r^{-N}} \]

We can use this formula to compute a value $r_i$ that produces the observed
value $x_i$, noting that for arbitrary strategies the value of this
effective relative fitness is dependent on $i$ and $N$.

@@

Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of the analytical formulae that relies on the average payoffs. A
detailed analysis of the 164 strategies considered, using direct Moran processes
will be shown in the next Section.

@@

% TODO Archive all data (Zenodo) and include details/reference here.

@@

\begin{itemize}
    \item Section~\ref{sec:two_individuals} considers the specific case of
        \(N=2\).
    \item Section~\ref{sec:strong_invaders} investigates the effect of
        population size on the ability of a strategy to invade another
        population. This will highlight how complex strategies with long
        memories outperform simpler strategies.
    \item Section~\ref{sec:strong_resistors} similarly investigates the
        ability to defend against an invasion.
    \item Section~\ref{sec:population_size} investigates the relationship
        between performance for differing population sizes.
\end{itemize}

@@

Overall the main fixation probabilities of interest are \(x_1\) and \(x_{N-1}\),
these reflect a strategy's ability to invade or resist invasion; for \(N=2\)
these two cases coincide. Figure~\ref{fig:boxplot_2} shows all fixation
probabilities for the strategies considered. This is summarised in
Table~\ref{tbl:summary_top_2}.

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism (a cooperation followed by a defection on the first
        move). As long as the opponent plays the same handshake and does not
        defect in the future it cooperates. Otherwise it defects for all
        rounds~\cite{Li2009}. This strategy was specifically designed for
        evolutionary processes.
    \item The Defector: it always defects. As it has little potential
        interaction with itself, recall that \(N=2\) is considered, its
        aggressiveness is rewarded.
    \item The Aggravater strategy which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in~\cite{Ashlock2006}.
    \item Handshake: a slightly less aggressive version of the Collective
        strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself: just as the Collective strategy does. However it
        is not aggressive enough to invade other populations.
\end{enumerate}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_2_invade.pdf}
    \caption{The fixation probabilities for \(N=2\)}
    \label{fig:boxplot_2}
\end{figure}

@@

As will be demonstrated in Section~\ref{sec:population_size} the results for
\(N=2\) differ from those of larger $N$. Hence our results do not concur with
the literature which suggests that Zero Determinant strategies should be
effective for larger population sizes, but we note that those analysis run each
match to stationarity, while our matches run for a fixed number of rounds.

In the next sections we pay close attention to
strategies who are strong invaders/resistors.

@@

In this section we focus on the ability of a mutant strategy to invade: the
probability of 1 individual of a given type successfully becoming fixated in a
population of \(N - 1\) other individuals, denoted by \(x_i\). The fixation
probabilities are shown in
Figures~\ref{fig:boxplot_3_invade},~\ref{fig:boxplot_7_invade}
and~\ref{fig:boxplot_14_invade} for \(N\in\{3, 7, 14\}\) showing the mean
fixation as well as the neutral fixation for each given scenario.

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_3_invade.pdf}
    \caption{The fixation probability \(x_1\) for \(N=3\)}
    \label{fig:boxplot_3_invade}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_7_invade.pdf}
    \caption{The fixation probability \(x_1\) for \(N=7\)}
    \label{fig:boxplot_7_invade}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_14_invade.pdf}
    \caption{The fixation probability \(x_1\) for \(N=14\)}
    \label{fig:boxplot_14_invade}
\end{figure}

@@

\begin{itemize}
    \item Grudger (which only performs well for \(N=3\)), starts by cooperating
        but will defect if at any point the opponent has defected.
    \item MEM2, an infinite memory strategy that switches between TfT, Tf2T, and
        Defector~\cite{Li2014}.
    \item TF1, the finite state machine trained specifically for Moran processes
        described in Section~\cite{sec:introduction}.
    \item Prober 4, complex strategy with an initial 20 move sequence of
        cooperations and defections~\cite{prison}. This initial sequence serves
        as approximate handshake.
    \item  PSO Gambler and Evolved Lookerup 2 2 2: are strategies that make use
        of a lookup table mapping the first 2 moves of the opponent as well as
        the last 2 moves of both players to an action. The PSO gambler is a
        stochastic version which maps those states to probabilities of
        cooperating. The lookerup was described in \cite{Knight2016}.
	\item The evolved ANN strategies are neural networks that map a number of
		attributes (first move, number of cooperations, last move etc\dots) to
		an action. Both of these have been trained using an evolutionary
		algorithm and the ANN 5 was trained to perform well in a noisy
		tournament.
    \item The Evolved FSM 16 is a 16 state finite state machine trained to
        perform well in tournaments.
\end{itemize}

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Only one is stochastic. The finite state machines trained
specifically for Moran processes do not appear in the top 5: whereas strategies
trained for tournament do. This is due to the nature of invasion: most of the
opponents will initially be different strategies.

The strategies trained with the mean score
objective are among the best invaders in the library but are not as resistant
to invasion as the strategies trained using a Moran objective function. These strategies include trained finite state machine
strategies, but they do not appear to have handshaking mechanisms. Therefore it
is reasonable to conclude that the objective function is the cause of the
emergence of handshaking mechanisms.
The payoff maximizing strategies
typically will not defect before the opponent's first defection, possibly
because the training strategy collection contains a significant portion of
strategies such as Grudger and Fool Me Once that retaliate harshly by defecting
for the remainder of the match if the opponent has more than a small number of
cumulative defections. Paradoxically it is necessary to defect in order to
achieve mutual cooperation with opponents using the same strategy but not with
other opponents. The importance of these handshakes will become apparent in the
next Section.

@@

In addition to identifying good invaders, we also identify strategies
resistant to invasion by other strategies by examining the distribution
of $x_{N-1}$ for each strategy. Note that this is equivalent to
looking at $x_1$ for all opponents.

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_3_resist.pdf}
    \caption{The fixation probability \(x_{N-1}\) for \(N=3\)}
    \label{fig:boxplot_3_resist}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.8\textheight]{./img/boxplot_7_resist.pdf}
    \caption{The fixation probability \(x_{N-1}\) for \(N=7\)}
    \label{fig:boxplot_6_resist}
\end{figure}

>>>>

% Title
\title{An Empirical Study of Invasion and Resistance
       for Iterated Prisoner's Dilemma Strategies}
\author{Vincent Knight \and
        Marc Harper \and
        Nikoleta E. Glynatsi \and
        Owen Campbell} % TODO Authors?
\date{}

@@

\begin{abstract}
    The Iterated Prisoner's Dilemma is a well established framework for
    the study of emergent behaviour. In this paper an extensive numerical
    study of the evolutionary dynamics of this framework are presented. Fixation
    probabilities for Moran processes are obtained for 164
    different strategies. It is found that players with long memories and
    sophisticated behaviours outperform many strategies that perform well
    in a two player setting.
\end{abstract}

@@

The Prisoner's Dilemma (PD) \cite{Flood1958} is a fundamental two player game
used to model a large variety of strategic interactions. Each player can choose
between cooperation (C) or defection (D). The decisions are made simultaneously
and independently. The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This extended form is called the Iterated Prisoner's
Dilemma (IPD). As described in \cite{Axelrod1980a, Knight2016, Press2012} a
number of strategies have been developed to take advantage of the history of
play. Recently, some strategies referred to as Zero Determinant strategies
\cite{Press2012} even manipulate players through extortion.

The Moran Process \cite{Moran1957} is a model of
evolutionary population dynamics that has been used to gain insights about the
evolutionary stability in a number of settings (more details given in
Section~\ref{sec:the_moran_process}).
Several earlier
works have studied iterated games in the context of the prisoner's dilemma
\cite{Nowak, Stewart2012}, however these often make simplifying assumptions
and/or do not consider sophisticated behaviour: only considering strategies that
either cooperate or defect.

This manuscript provides a detailed numerical analysis of
\textbf{\input{./tex/num_strategies.tex}}complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library \cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over \input{./tex/num_strategies_axelrod.tex}parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, Zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: same use
knowledge of the game such as the number of turns and others have a high
computational run time. The large number of strategies are available thanks to
the open source nature of the project with over 40 contributions made by
different programmers and researchers \cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in Section~\ref{sec:strategies}.

The library can conduct matches, tournaments and population
dynamics with variations including noise and spatial structure.
The strategies and simulation frameworks are
automatically tested to an extraordinarily high degree of coverage in accordance
with best research software practices.

Fixation probabilities for all pairs of
strategies are presented, identifying those that are effective invaders and
those resistant to invasion, for population sizes $N=2$ to $N=14$.

In particular the following questions are addressed:
\begin{enumerate}
    \item What strategies are good invaders?
    \item What strategies are good at resisting invasion?
    \item How does the population size affect these findings?
\end{enumerate}

While the results agree with some of the published literature, it is found that:

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform intelligently designed
     strategies.
 \item Strong resistors specifically evolve or have a handshake mechanism. This
     has the potential to offer insight in to organisms such as antibacterial
     resistant bacteria \cite{Davies2010}.
\end{enumerate}

\subsection{The Moran Process}\label{sec:the_moran_process}

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process, a stochastic birth death process on a finite population in which the
population size stays constant over time. Individuals are \textbf{selected}
according to a given fitness landscape. Once selected, the individual is
reproduced and similarly another individual is chosen to be removed from the
population. In some settings mutation is also considered but without mutation
(the case considered in this work) this process will arrive at an absorbing
state where the population is entirely made up of players of one strategy. The
probability with which a given strategy is the survivor is called the
\textit{fixation probability}. A more detailed analytic description of this is
given in Section~\ref{sec:validation}.

@@

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider non-sophisticated strategies. Some work has
looked at evolutionary stability of strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies, strategies that recall
the events of the previous round only.

Of particular interest are the Zero determinant strategies introduced in
\cite{Press2012} and praised in \cite{Stewart2012} it was argued that generous
ZD strategies are robust against invading strategies. However, in \cite{Lee2015}
a strategy using machine learning techniques was capable of resisting invasion
and also able to invade any memory one strategy. Recent work \cite{Hilbe2017}
has investigated the effect of memory length on strategy performance and the
emergence of cooperation but this is not done in Moran process context and only
considers specific cases of memory 2 strategies. In \cite{Adami2013} it was
recognised that many Zero determinant strategies do not fare well against
themselves. This is a disadvantage for the Moran process where the best
strategies cooperate well with other players using the same strategy.

\subsection{Strategies considered}\label{sec:strategies}

To carry out this large numerical experiment, \input{./tex/num_strategies}
strategies, listed in Appendix~\ref{app:list_of_players} are used library. There
are \input{./tex/num_stochastic.tex}stochastic and
\input{./tex/num_deterministic.tex}deterministic strategies. Their memory depth,
defined by the number of rounds of history used by the strategy each round, is
shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if the
strategy uses the entire history of play (whatever its length). For example, a
strategy that utilizes a handshaking mechanism where the opponents actions on
the first few rounds of play determines the strategies subsequent behavior would
have infinite memory depth.

A number of these strategies have been trained with reinforcement learning
algorithms.

\begin{itemize}
    \item Evolved ANN: a neural network based strategy;
    \item Evolved LookerUp: a lookup table based strategy;
    \item PSO Gambler: a stochastic version of the lookup table based strategy;
    \item Evolved HMM: a hidden Markov model based strategy.
\end{itemize}

A part from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, all these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.

More
information about each player can be obtained in the documentation for
\cite{axelrodproject} and a detailed description of the performance
 of these strategies in IPD
tournaments will be described in upcoming manuscript.
% TODO Think of how to reference this upcoming manuscript

All of the training code is available in the
Axelrod repository with documentation to train new strategies easily. Training
% TODO Include link etc
typically takes less than 200 iterations and can be completed within several
hours on commodity hardware.

There are three further strategies trained specifically for this study; Trained
FSM 1, 2, and 3 (TF1 - TF3). These are based on finite state machines of 8, 16,
and 16 states respectively (see Figures \ref{}).
% TODO Include images of FSMs
As opposed to the previously described strategies, these strategies were trained
with the objective function of \textbf{mean fixation probabilities for Moran
processes} starting at initial population states consisting of \(N/2\)
individuals of the training candidates and \(N/2\) individuals of an opponent
strategy, taken from a selection of 150 opponents from the axelrod library:

@@

Each matchup of players was run to fixation for the specified number of
repetitions to estimate the absorption probabilities. The trained algorithms
were run for less than 50 generations. Training data for this is available at
\cite{data}.

TF1 cooperates and defects with
various cycles depending on the opponent's actions. TF1 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective. Interestingly it has no handshake
mechanism, this could be due to the fact that it is trained in a noisy
environment.

TF2 always starts with CD and will defect against opponents that start with
DD\@. It plays CDD against itself and then cooperates thereafter. There is a
longer complex handshake which eventually results in mutual cooperation with
Firm but Fair, Fortress3, Fortress4, and Grofman (always) and Evolved HMM 5 and
GTFT (depending on the random seed).

@@

For both TF2 and TF3 a handshake
mechanism naturally emerges from the structure of the underlying finite state
machine. This behavior is an outcome of the evolutionary process and is in no
way hard-coded or included via an additional mechanism.

@@

\subsection{Data collection}\label{sec:data_collection}

Each strategy pair is run for 1000 repetitions of the Moran process to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken
from a large number of match outcomes for the pairs of players for use in
computing fitnesses in the Moran process. This approach was verified to agree
with unsampled calculations to a high degree of accuracy in specific cases.
This is described in Figure~\ref{alg:psuedocode}.
% TODO Include pseudo code

Section~\ref{sec:validation} will further validate the methodology by comparing
simulated results to analytical results in some cases. The main results of this
manuscript are presented in Section~\ref{sec:empirical_results} which will
present a detailed analysis of all the data generated. Finally,
Section~\ref{sec:conclusion} will conclude and offer future avenues for the work
presented here.

@@

Using this it is a known result \cite{Nowak2017} that the fixation probability
of the first strategy in a population of \(i\) individuals of the first type
(and \(N-i\) individuals of the second::

@@

A neutral strategy will have fixation probability $x_i = i/N$.

@@

Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of the analytical formulae that relies on the average payoffs
(\ref{equ:payoff_matrix}).

All data generated for this validation exercise can be found
at \cite{data}.

A detailed analysis of the 164 strategies considered,
using direct Moran processes will be shown in the next Section.

@@

@@

\begin{itemize}
    \item Section~\ref{sec:two_individuals} considers the specific case of
        \(N=2\).
    \item Section~\ref{sec:strong_invaders} investigates the effect of
        population size on the ability of a strategy to invade another
        population. This will highlight how complex strategies with long
        memories outperform simpler strategies.
    \item Section~\ref{sec:strong_resistors} similarly investigates the
        ability to defend against an invasion.
    \item Section~\ref{sec:population_size} investigates the relationship
        between performance for differing population sizes as well as
        taking a close look at Zero determinant strategies \cite{Press2012}.
\end{itemize}

@@

For \(N=2\) the two cases of \(x_1\) and \(x_{N-1}\) coincide, but will be
considered separately for larger \(N\) in sections~\ref{sec:strong_invaders} and
~\ref{sec:strong_resistors}.  Figure~\ref{fig:boxplot_2} shows all fixation
probabilities for the strategies considered. This is summarised in
Table~\ref{tbl:summary_top_2}.

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism (a cooperation followed by a defection on the first
        move). As long as the opponent plays the same handshake and does not
        defect in the future it cooperates. Otherwise it defects for all
        rounds. This strategy was specifically designed for
        evolutionary processes \cite{Li2009}.
    \item The Defector: it always defects. As it has little potential
        interaction with itself (recall that \(N=2\)), its
        aggressiveness is rewarded.
    \item The Aggravater strategy which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in \cite{Ashlock2006}.
    \item Handshake: a slightly less aggressive version of the Collective
        strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself: just as the Collective strategy does. However it
        is not aggressive enough to invade other populations (as can be seen in
        Section~\ref{sec:strong_invaders}.
\end{enumerate}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/boxplot_2_invade.pdf}
    \caption{The fixation probabilities for \(N=2\)}
    \label{fig:boxplot_2}
\end{figure}

@@

As will be demonstrated in Section~\ref{sec:population_size} the results for
\(N=2\) differ from those of larger $N$. Hence these results do not concur with
the literature which suggests that Zero Determinant strategies should be
effective for larger population sizes, but these analysis consider stationary
behaviour, while this work runs for a fixed number of rounds. Note that the
stationarity assumptions allows for the analysis to take place leading to the
conclusions about zero determinant strategies however it is dependent on a
specific structure of strategies. The analysis carried out here makes no
assumptions about the structure of the strategies by using actual simulation
interactions.

In the next sections close attention to
strategies who are strong invaders/resistors is given.

@@

In this section the focus is on the ability of a mutant strategy to invade: the
probability of 1 individual of a given type successfully becoming fixated in a
population of \(N - 1\) other individuals, denoted by \(x_1\). The fixation
probabilities are shown in
Figures~\ref{fig:boxplot_3_invade},~\ref{fig:boxplot_7_invade}
and~\ref{fig:boxplot_14_invade} for \(N\in\{3, 7, 14\}\) showing the mean
fixation as well as the neutral fixation for each given scenario.

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/boxplot_3_invade.pdf}
    \caption{The fixation probabilities \(x_1\) for \(N=3\)}
    \label{fig:boxplot_3_invade}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/boxplot_7_invade.pdf}
    \caption{The fixation probabilities \(x_1\) for \(N=7\)}
    \label{fig:boxplot_7_invade}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/boxplot_14_invade.pdf}
    \caption{The fixation probabilities \(x_1\) for \(N=14\)}
    \label{fig:boxplot_14_invade}
\end{figure}

@@

\begin{itemize}
    \item Grudger (which only performs well for \(N=3\)), starts by cooperating
        but will defect if at any point the opponent has defected.
    \item MEM2, an infinite memory strategy that switches between TfT, Tf2T, and
        Defector \cite{Li2014}.
    \item TF1, the finite state machine trained specifically for Moran processes
        described in Section~\ref{sec:introduction}.
    \item Prober 4, complex strategy with an initial 20 move sequence of
        cooperations and defections \cite{prison}. This initial sequence serves
        as approximate handshake.
    \item  PSO Gambler and Evolved Lookerup 2 2 2: are strategies that make use
        of a lookup table mapping the first 2 moves of the opponent as well as
        the last 2 moves of both players to an action. The PSO gambler is a
        stochastic version which maps those states to probabilities of
        cooperating. The lookerup was described in \cite{Knight2016}.
	\item The evolved ANN strategies are neural networks that map a number of
		attributes (first move, number of cooperations, last move etc\dots) to
		an action. Both of these have been trained using an evolutionary
		algorithm and the ANN 5 was trained to perform well in a noisy
		tournament.
    \item The Evolved FSM 16 is a 16 state finite state machine trained to
        perform well in tournaments.
\end{itemize}

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Only one is stochastic although close inspection of the
source code of PSO Gambler shows that it makes stochastic decisions with extreme
probabilities: almost always acting as expected. Apart from TF1 in \(N=3\), the
finite state machines trained specifically for Moran processes do not appear in
the top 5: whereas strategies trained for tournament do. This is due to the
nature of invasion: most of the opponents will initially be different
strategies. The next section will consider the converse situation.

@@

In addition to identifying good invaders, strategies resistant to invasion by
other strategies are identified by examining the distribution of $x_{N-1}$ for
each strategy. Note that this is equivalent to looking at $x_1$ for all
opponents.

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/boxplot_3_resist.pdf}
    \caption{The fixation probability \(x_{N-1}\) for \(N=3\)}
    \label{fig:boxplot_3_resist}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/boxplot_7_resist.pdf}
    \caption{The fixation probability \(x_{N-1}\) for \(N=7\)}
    \label{fig:boxplot_6_resist}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/boxplot_14_resist.pdf}
    \caption{The fixation probability \(x_{N-1}\) for \(N=14\)}
    \label{fig:boxplot_14_resist}
\end{figure}
